More straightforward movement variable definitions

### 2024-02-14 scripts/01.1_categorizeMovement.qmd
updated with new space use PC

# Create a dataset with all seasons combined
allMovementBehavior <- purrr::list_rbind(movementBehavior)
movementBehaviorScaled <- allMovementBehavior %>%
  mutate(across(-c(Nili_id, seasonUnique, birth_year, sex, homeRange, coreArea, coreAreaFidelity), # not scaling the space use params, because I'm not putting those into a PCA.
                function(x){as.numeric(as.vector(scale(x)))}))
```

@@

```{r}
#| include: false
demo <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, daysTracked, propDaysTracked, birth_year, sex)
space_use <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, coreArea, homeRange, coreAreaFidelity)
movement <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, mnDailyMnAlt, meanDFD, meanDMD, meanDDT, mnTort)
roost_behavior <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, propSwitch, shannon, uniqueRoosts)
```

@@

@@

```{r}
#| message: false
#| warning: false
#| include: false
# Create a reduced dataset that only includes our measures of interest
red <- demo %>%
  left_join(space_use %>% dplyr::select(Nili_id, seasonUnique, homeRange)) %>%
  left_join(movement %>% dplyr::select(Nili_id, seasonUnique, movement_pc1)) %>%
  left_join(roost_behavior %>% dplyr::select(Nili_id, seasonUnique, roost_pc1)) %>%
  rename("space_use" = "homeRange",
         "movement" = "movement_pc1",
         "roost_div" = "roost_pc1")
nrow(red) == nrow(demo) # checking that the number of rows hasn't changed
red_long <- red %>%
  pivot_longer(cols = c("space_use", "movement", "roost_div"),
               names_to = "var",
               values_to = "value")
```

Let's examine these predictor variables

@@

We can see here that we have another problem. Since `movement` and `roost_div` are composite variables derived from PCAs, they are pretty close to normally distributed. Meanwhile, our space use variable is very right-skewed, because it's a measure of area.

Let's log-tranform the space use variable to make the distributions more comparable and therefore make the modeling easier.

```{r}
#| include: false
red <- red %>%
  mutate(space_use_log = log(space_use))
red_long <- red %>%
  pivot_longer(cols = c("space_use", 
                        "space_use_log", 
                        "movement", 
                        "roost_div"),
               names_to = "var",
               values_to = "value")
```

Re-examine the distributions of the variables after log-transforming:

```{r}
#| message: false
#| warning: false
red_long %>%
  filter(var != "space_use") %>%
  ggplot(aes(x = value, col = var, fill = var))+
  geom_histogram()+ theme_minimal()+
  facet_wrap(~var, scales = "free")+
  scale_fill_manual(values = colors)+
  scale_color_manual(values = colors)+
  theme(legend.position = "none")+
  ylab("Frequency")+xlab("")
```

Okay that's still not perfect, but it's a lot better.

*Note also that this outcome may change quantitatively after we calculate the home ranges by better means (AKDE), but I don't expect it to be qualitatively different. AKDE calculations are currently in progress.*

@@

**Roost diversification** and **log-transformed home range** have a stronger correlation of `r round(cor(red$roost_div, red$space_use_log), 2)`.

```{r}
#| message: false
#| warning: false
cor <- round(cor(red$roost_div, red$space_use_log), 2)
red %>%
  ggplot(aes(x = roost_div, y = space_use_log))+
  geom_point(color = "chartreuse3")+ geom_smooth(method = "lm", col = "chartreuse3")+ theme_minimal()+
  xlab("Roost diversification")+ ylab("Space use (log)")+
  ggplot2::annotate(geom = "text", x = 4, y = 11, label = paste0("Correlation = ", cor))
```

@@

```{r}
#| message: false
#| warning: false
cor <- round(cor(red$space_use_log, red$movement), 2)
red %>%
  ggplot(aes(x = space_use_log, y = movement))+
  geom_point(color = "orange2")+ geom_smooth(method = "lm", col = "orange2")+ theme_minimal()+
  xlab("Space use (log)") + ylab("Movement")+
  ggplot2::annotate(geom = "text", x = 14, y = -4.5, label = paste0("Correlation = ", cor))
```

@@

```{r}
#| warning: false
#| message: false
## TEMPORARY: FOR PRESENTATION
load(here::here("data", "derived", "cc.Rda"))
red %>%
  mutate(season = stringr::str_extract(seasonUnique, "[a-z]+")) %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use_log"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = age, y = value, group = seasonUnique, col = season))+
  geom_point(alpha = 0.5) + geom_smooth(se = F)+
  theme_minimal()+
  xlab("Age")+
  facet_wrap(~var, scales = "free")+
  scale_color_manual(values = c(cc$breedingColor, cc$fallColor, cc$summerColor))+
  theme()
```

```{r}
#| warning: false
#| message: false
red %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use_log"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = age, y = value, group = seasonUnique, col = var))+
  geom_point(alpha = 0.5) + geom_smooth(se = F)+
  theme_minimal()+
  xlab("Age")+
  facet_wrap(~var, scales = "free")+
  scale_color_manual(values = colors)+
  theme(legend.position = "none")
```

@@

```{r}
#| warning: false
#| message: false
red %>%
  filter(!is.na(sex), sex != "NA") %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use_log"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = sex, y = value, col = var))+
  geom_jitter(height = 0, width = 0.2, alpha = 0.7)+
  geom_violin(fill = NA)+
  theme_minimal()+
  xlab("Age")+
  facet_wrap(~var, scales = "free")+
  scale_color_manual(values = colors)+
  theme(legend.position = "none")+ylab("")
# 
# red %>%
#   filter(!is.na(sex), sex != "NA") %>%
#   pivot_longer(cols = c("movement", "roost_div", "space_use_log"),
#                names_to = "var", values_to = "value") %>%
#   ggplot(aes(x = sex, y = value, col = var))+
#   geom_jitter(height = 0, width = 0.2, alpha = 0.7)+
#   geom_violin(fill = NA)+
#   theme_minimal()+
#   xlab("Age")+
#   facet_grid(rows = vars(var), cols = vars(season), scales = "free")+
#   scale_color_manual(values = colors)+
#   theme(legend.position = "none")+ylab("")
```

@@

```{r}
red %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use_log"),
               names_to = "var", values_to = "value") %>%
  filter(var != "space_use") %>%
  ggplot(aes(x = value, group = seasonUnique, col = season))+
  geom_density()+ theme_minimal()+
  scale_color_manual(name = "Season", 
                     values = c(cc$breedingColor, 
                                cc$summerColor, 
                                cc$fallColor))+
  facet_wrap(~var, scales = "free")+
  ylab("Frequency")+xlab("")

red %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use_log"),
               names_to = "var", values_to = "value") %>%
  filter(var != "space_use") %>%
  ggplot(aes(x = season, y = value))+
  geom_violin(aes(fill = season, col = season))+
  facet_wrap(~var, scales = "free")+
  theme_minimal()+
  scale_color_manual(name = "Season", 
                     values = c(cc$breedingColor, 
                                cc$summerColor, 
                                cc$fallColor))+
  scale_fill_manual(name = "Season", 
                     values = c(cc$breedingColor, 
                                cc$summerColor, 
                                cc$fallColor))+
  theme(legend.position = "none")+
  ylab("Value")+
  xlab("Season")
```

>>>>

# Create a dataset with all seasons combined
allMovementBehavior <- purrr::list_rbind(movementBehavior)
movementBehaviorScaled <- allMovementBehavior %>%
  mutate(coreArea_log = log(coreArea),
         homeRange_log = log(homeRange)) %>%
  mutate(across(-c(Nili_id, seasonUnique, birth_year, sex), 
                function(x){as.numeric(as.vector(scale(x)))}))
```

@@

```{r}
#| include: false
demo <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, daysTracked, propDaysTracked, birth_year, sex)

space_use <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, coreArea_log, homeRange_log, coreAreaFidelity)

movement <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, mnDailyMnAlt, meanDFD, meanDMD, meanDDT, mnTort) %>%
  filter(!is.na(mnDailyMnAlt))

roost_behavior <- movementBehaviorScaled %>%
  dplyr::select(Nili_id, seasonUnique, propSwitch, shannon, uniqueRoosts) %>%
  filter(!is.na(shannon))
```

@@

## Space use PCA

```{r}
pca_space <- prcomp(space_use[,-c(1:3)]) # only including home range and coreAreaFidelity (because it would be redundant to include all three)
fviz_pca_biplot(pca_space, axes = c(1,2), 
                geom = c("point"), 
                alpha.ind = 0.3,
                repel = T)+
  theme_classic()
fviz_screeplot(pca_space, addLabels = T) # This is also strongly loaded on one principal component, so I feel quite comfortable taking it as a measure of roost diversification/routine
contrib_space <- round(pca_roost$rotation[,1:3]*-100, 2) %>%
  as.data.frame()
contrib_space
space_use <- space_use %>%
  mutate(space_pc1 = pca_space$x[,1]*-1)
```

@@

```{r}
#| message: false
#| warning: false
#| include: false
# Create a reduced dataset that only includes our measures of interest
red <- demo %>%
  left_join(space_use %>% dplyr::select(Nili_id, seasonUnique, space_pc1)) %>%
  left_join(movement %>% dplyr::select(Nili_id, seasonUnique, movement_pc1)) %>%
  left_join(roost_behavior %>% dplyr::select(Nili_id, seasonUnique, roost_pc1)) %>%
  rename("space_use" = "space_pc1",
         "movement" = "movement_pc1",
         "roost_div" = "roost_pc1") %>%
  filter(!is.na(movement))
nrow(red) == nrow(movement) # checking that the number of rows hasn't changed
red_long <- red %>%
  pivot_longer(cols = c("space_use", "movement", "roost_div"),
               names_to = "var",
               values_to = "value")
```

Let's examine these predictor variables.

@@

All variables are fairly close to normally distributed, since they come from PCAs and are scaled.

@@

**Roost diversification** and **log-transformed home range** have a stronger correlation of `r round(cor(red$roost_div, red$space_use), 2)`.

```{r}
#| message: false
#| warning: false
cor <- round(cor(red$roost_div, red$space_use), 2)
red %>%
  ggplot(aes(x = roost_div, y = space_use))+
  geom_point(color = "chartreuse3")+ geom_smooth(method = "lm", col = "chartreuse3")+ theme_minimal()+
  xlab("Roost diversification")+ ylab("Space use (log)")+
  ggplot2::annotate(geom = "text", x = 3, y = -2, label = paste0("Correlation = ", cor))
```

@@

```{r}
#| message: false
#| warning: false
cor <- round(cor(red$space_use, red$movement), 2)
red %>%
  ggplot(aes(x = space_use, y = movement))+
  geom_point(color = "orange2")+ geom_smooth(method = "lm", col = "orange2")+ theme_minimal()+
  xlab("Space use (log)") + ylab("Movement")+
  ggplot2::annotate(geom = "text", x = 2.5, y = -3, label = paste0("Correlation = ", cor))
```

@@

```{r}
#| warning: false
#| message: false
## TEMPORARY: FOR PRESENTATION
load(here::here("data", "derived", "cc.Rda"))
red %>%
  mutate(season = stringr::str_extract(seasonUnique, "[a-z]+")) %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = age, y = value, group = seasonUnique, col = season))+
  geom_point(alpha = 0.5) + geom_smooth(se = F)+
  theme_minimal()+
  xlab("Age")+
  facet_wrap(~var, scales = "free")+
  scale_color_manual(values = c(cc$breedingColor, cc$fallColor, cc$summerColor))+
  theme()
```

```{r}
#| warning: false
#| message: false
red %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = age, y = value, group = seasonUnique, col = var))+
  geom_point(alpha = 0.5) + geom_smooth(se = F)+
  theme_minimal()+
  xlab("Age")+
  facet_wrap(~var, scales = "free")+
  scale_color_manual(values = colors)+
  theme(legend.position = "none")
```

@@

```{r}
#| warning: false
#| message: false
red %>%
  filter(!is.na(sex), sex != "NA") %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = sex, y = value, col = var))+
  geom_jitter(height = 0, width = 0.2, alpha = 0.7)+
  geom_violin(fill = NA)+
  theme_minimal()+
  xlab("Age")+
  facet_wrap(~var, scales = "free")+
  scale_color_manual(values = colors)+
  theme(legend.position = "none")+ylab("")
# 
# red %>%
#   filter(!is.na(sex), sex != "NA") %>%
#   pivot_longer(cols = c("movement", "roost_div", "space_use"),
#                names_to = "var", values_to = "value") %>%
#   ggplot(aes(x = sex, y = value, col = var))+
#   geom_jitter(height = 0, width = 0.2, alpha = 0.7)+
#   geom_violin(fill = NA)+
#   theme_minimal()+
#   xlab("Age")+
#   facet_grid(rows = vars(var), cols = vars(season), scales = "free")+
#   scale_color_manual(values = colors)+
#   theme(legend.position = "none")+ylab("")
```

@@

```{r}
red %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use"),
               names_to = "var", values_to = "value") %>%
  filter(var != "space_use") %>%
  ggplot(aes(x = value, group = seasonUnique, col = season))+
  geom_density()+ theme_minimal()+
  scale_color_manual(name = "Season", 
                     values = c(cc$breedingColor, 
                                cc$summerColor, 
                                cc$fallColor))+
  facet_wrap(~var, scales = "free")+
  ylab("Frequency")+xlab("")

red %>%
  pivot_longer(cols = c("movement", "roost_div", "space_use"),
               names_to = "var", values_to = "value") %>%
  ggplot(aes(x = season, y = value))+
  geom_violin(aes(fill = season, col = season))+
  facet_wrap(~var, scales = "free")+
  theme_minimal()+
  scale_color_manual(name = "Season", 
                     values = c(cc$breedingColor, 
                                cc$summerColor, 
                                cc$fallColor))+
  scale_fill_manual(name = "Season", 
                     values = c(cc$breedingColor, 
                                cc$summerColor, 
                                cc$fallColor))+
  theme(legend.position = "none")+
  ylab("Value")+
  xlab("Season")
```
